Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2023-06-19 unimodular-covers-CombTheory-final.tex

\section{Proof of \Cref{lemma:cayley}}
\label{sec:the_lemma}
Let $f_q$ be the primitive lattice functional constant on $q$ and $f_p$ the one constant on $p$. We assume that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.

Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that $q_1$ is the origin and that $f_q(p_1)=-1$. 
Observe that then $f_q(p_2)$ must be strictly positive, since $f_q(p_2)=0$ would imply that $p+q$ is a unimodular paralellogram. Moreover, since $p$ is primitive $p_1$ and $p_2$ cannot be in the boundary of $q +\vecline p$, which implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

\begin{proof}
Both $t$ and $b$ must cross the interior of $p$, or otherwise $p_1$ or $p_2$ are the lattice points we are looking for in $Q$.
To seek a contradiction assume, as in \Cref{fig:claim2}, that 
\[
f_q(t_2 - t_1) \le 0 \le f_q(b_2-b_1).
\]
That is, $f_q$ decreases (perhaps weakly) along $t$ and increases along $b$, as $f_p$ increases on both. This implies that $Q \cap V_2$ is contained in the open strip $\{f_q(p_1) <f_q(x)< f_q(p_2)\}$, of width $w$.
%
%
%Then, $f_q(t_2) \le  f_q (t_1)$ implies $f_q(t_2)$ equals the minimum of $f_q$ on $t$, which is smaller than $f_q(p_2)=w-1$. Similarly, 
%
% f_q(t_2)
%
%This implies 
%
%
%
%Let us first show that $f_q(b_2-b_1)\ne 0 \ne f_q(t_2 - t_1)$. If, say, $f_q(b_2-b_1)= 0$, then $b$ is parallel to $q$ and its endpoints $b_1$ and $b_2$ must be on the boundary of the strip. By \Cref{claim:width} this implies $b=q$.
%
%Then, $[b_i,q_i]$ is parallel to $p$, for $i=1,2$.
%
%The left (or right) endpoint of this translate and of $q$ are lattice points lying on a line parallel to $p$, and since both this translate of $q$ and $q$ cross the segment $p$, they force $p$ to contain a lattice point in the interior, a contradiction to $p$ being primitive.  
%	
%Suppose by contradiction that the maximum of $f_q$ on $t$  lies in $V_1$ and that the maximum on $b$ lies in $V_2$. %
%Then $Q \cap V_2$ is contained in the open strip $\{-1<f_q(x)<w-1\}$, of width $w$. 
This strip cannot contain a translated copy of $r$, since $\width_{f_q}(r) \geq w$, which gives a contradiction: since $P$ is a weak Minkowski summand of $Q$,  $Q$ must have edges parallel to $r$ both in $V_1$ and $V_2$.
\end{proof}

@@

We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must have an edge $r'$ which is a translated copy of $r$. Let $r_1'$ and $r_2'$ be its endpoints.

Consider the segment $s$ contained in $r_1' + \vecline p$ with endpoints $s_1=r_1'$ on $d$ and $s_2$ on the line spanned by $t$. The width of $s$ with respect to $g$ is $g(s_2)-g(s_1) <1$, because $s$ is shorter than $Q \cap p$, which is strictly contained between the consecutive parallel lattice lines through $p_1$ and $q_2$ and through $q_1$ and $q_1+q_2-p_1$ where $g$ is constant.

We now observe that $g(r_2) < g(s_2)$, because $r_2$ lies to the right of $s_2$ ($r$ has positive slope) and below the line spanned by $t$, and $g$ decreases moving to the right along $t$ (\Cref{claim:g}). Thus $g(r_2)-g(r_1) = g(r'_2)-g(r'_1) < g(s_2)-g(s_1) <1$, that is, $g(r_2)=g(r_1)$. This contradicts \Cref{claim:g}.
%Since $t$ separates $q_1$ and $q_1+q_2-p_1$ and $g$ decreases from $t_1$ to $t_2$ (by \Cref{claim:g}), the inequality $g(x)< g(d')$ holds on $Q\cap V_2$, and in particular for $r_2'$. Since $r_2'$ is a lattice point, $g(r_2')\leq g(d)= g(r_1')$, which contradicts \Cref{claim:g}).
\end{proof}

>>>>

\section{Proof of \Cref{lemma:cayley}}
\label{sec:the_lemma}
Let $f_q$ be the primitive lattice functional constant on $q$ and $f_p$ the one constant on $p$. We assume that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.
Observe that since the segments $p_1p_2$ and $q_1q_2$ are primitive, the fact the $f_p$ and $f_q$ are also primitive implies that the area of the parallelogram $p+q$ equals both $f_p(q_2) - f_p(q_1)$ and $f_q(p_2) - f_q(p_1)$.
%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that $q_1$ is the origin, that $p$ belongs to the strip $q+\vecline p$  and that $f_q(p_1)=-1$.

Then $f_q(p_2)$ must be strictly positive, since $f_q(p_2)=0$ would imply that $p+q$ is a unimodular paralellogram. Moreover, since $p$ is primitive  $p$ must be in the interior of $q +\vecline p$ or otherwise we would have $p_2$ equal to one of $q_1$ or $q_2$, and then $f_q(p_2)=0$.
This implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

\begin{proof}
Both $t$ and $b$ must cross the interior of $p$, or otherwise $p_1$ or $p_2$ are the lattice points we are looking for in $Q$.
To seek a contradiction assume, as in \Cref{fig:claim2}, that 
\[
f_q(t_2 - t_1) \le 0 \le f_q(b_2-b_1).
\]
That is, $f_q$ decreases (perhaps weakly) along $t$ and increases along $b$, as $f_p$ increases on both. This implies that $Q \cap V_2$ is contained in the open strip $\{f_q(p_1) <f_q(x)< f_q(p_2)\}$, of width $w$.
%
%
%Then, $f_q(t_2) \le  f_q (t_1)$ implies $f_q(t_2)$ equals the minimum of $f_q$ on $t$, which is smaller than $f_q(p_2)=w-1$. Similarly, 
%
% f_q(t_2)
%
%This implies 
%
%
%
%Let us first show that $f_q(b_2-b_1)\ne 0 \ne f_q(t_2 - t_1)$. If, say, $f_q(b_2-b_1)= 0$, then $b$ is parallel to $q$ and its endpoints $b_1$ and $b_2$ must be on the boundary of the strip. By \Cref{claim:width} this implies $b=q$.
%
%Then, $[b_i,q_i]$ is parallel to $p$, for $i=1,2$.
%
%The left (or right) endpoint of this translate and of $q$ are lattice points lying on a line parallel to $p$, and since both this translate of $q$ and $q$ cross the segment $p$, they force $p$ to contain a lattice point in the interior, a contradiction to $p$ being primitive.  
%	
%Suppose by contradiction that the maximum of $f_q$ on $t$  lies in $V_1$ and that the maximum on $b$ lies in $V_2$. %
%Then $Q \cap V_2$ is contained in the open strip $\{-1<f_q(x)<w-1\}$, of width $w$. 
This strip cannot contain a translated copy of $r$, since $\width_{f_q}(r) \geq w$, which gives a contradiction: since $P$ is a weak Minkowski summand of $Q$, $Q$ must have an edge parallel to $r$ and with exterior normal pointing to the right. This edge must lie in $V_2$, which is impossible.
\end{proof}

@@

We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must contain an edge parallel to $r$ on the right. Let $r'$ be a primitive segment along this edge, so that $r'$ is a translated copy of $r$.
 Let $r_1'$ and $r_2'$ be its endpoints.

Consider the segment $s$ contained in $r_1' + \vecline p$ with endpoints $s_1=r_1'$ and $s_2$ on the line spanned by $t$. The width of $s$ with respect to $g$ is $g(s_2)-g(s_1) <1$, because $s$ is shorter than $Q \cap p$, which is strictly contained between the consecutive parallel lattice lines through $p_1$ and $q_2$ and through $q_1$ and $q_1+q_2-p_1$ where $g$ is constant.

We now observe that $g(r'_2) < g(s_2)$, because $r'_2$ lies to the right of $s_2$ ($r$ has positive slope) and below the line spanned by $t$, and $g$ decreases moving to the right along $t$ (\Cref{claim:g}). Thus $g(r_2)-g(r_1) = g(r'_2)-g(r'_1) < g(s_2)-g(s_1) <1$, that is, $g(r_2)=g(r_1)$. This contradicts \Cref{claim:g}.
%Since $t$ separates $q_1$ and $q_1+q_2-p_1$ and $g$ decreases from $t_1$ to $t_2$ (by \Cref{claim:g}), the inequality $g(x)< g(d')$ holds on $Q\cap V_2$, and in particular for $r_2'$. Since $r_2'$ is a lattice point, $g(r_2')\leq g(d)= g(r_1')$, which contradicts \Cref{claim:g}).
\end{proof}
